---
title: Predicting Fuel Prices and Impact Level Based on Brent Price and Country's
  Characteristics
author: "Group 2"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Volatilities in global energy markets introduce severe economic uncertainties, impacting consumer retail prices and national economic planning. Understanding the pass-through effect of crude oil price shocks onto localized retail markets requires a dual-pronged analytical framework.

This study leverages historical data across distinct temporal granularities to address two core research questions using machine learning paradigms:

#### Q1: Retail Price Prediction (Regression Model)
**Objective:** To predict future retail fuel prices based on the historical behavior of crude oil benchmarks.

**Granularity:** Analyzed according to country by month to capture medium-term macroeconomic shifts and regional dependencies.

**Approach:** Continuous numeric forecasting (Regression) to map upstream supply-side costs directly to downstream consumer impacts.

#### Q2: Global Price Impact Assessment (Classification Model)
**Objective:** To build a classification architecture that analyzes the specific impact level of global fuel price spikes on localized retail adjustments.

**Granularity:** Monitored by quarter to align with broader corporate and national budgetary cycles.

**Approach:** Categorical risk assessment (Classification) to segment country-specific fuel behavior into distinct shock-absorption or impact bands.

We will employ a standard machine learning framework which consist of:

- **Data Understanding:** Overview and surface understanding of the data to ensure it meets our needs
- **Data Preprocessing:** Cleaning, integration and transformation to ensure the data is robust
- **Exploratory Data Analysis:** Exploring the data deeper to uncover relationships through visualization
- **Modeling:** Compare various models along with hyperparameter tuning to find the best suited for both questions
- **Analysis:** Analyze the result, as well as understand the limitation

---

## Data Understanding

The data was gathered from Kaggle by the user BELBIN BENO R M titled [*Global Fuel Prices 2020–2026*](https://www.kaggle.com/datasets/belbino/global-fuel-prices-20202026). It was then imported to R with the default function. A preview of the first 5 records are as below:

``` {r data-import}
df <- read.csv("./data_raw/global_fuel_prices_2020_2026.csv")
```
``` {r data-preview, echo=FALSE}
library(knitr)
kable(df[1:5, ], caption = "Preview of the First Five Records in the Dataset")
```

The data consists of nine features:

- `date`: The time stamp of the record in the format of YYYY-MM-DD
- `country`: The specific nation where the fuel prices were recorded
- `region`: The broader geographic territory
- `income_level`: The economic classification of the country
- `subsidy_level`: Indicates the scale of government intervention and subsidies on national fuel prices
- `petrol_usd_liter`: The local retail price of consumer gasoline, standardized to US Dollars per liter
- `diesel_usd_liter`: The local retail price of automotive and industrial diesel fuel, standardized to US Dollars per liter
- `lpg_usd_liter`: The local retail price of Liquefied Petroleum Gas, standardized to US Dollars per liter
- `brent_crude_usd`: The global benchmark price for a barrel of Brent Crude oil in US Dollars
- `tax_percentage`: The tax rate applied by the national government to fuel sales

``` {r preview-again, echo=FALSE, include = FALSE}
library(dplyr)
library(tidyverse)

# 1. Identify categorical columns (Factors or Characters)
categorical_cols <- df %>%
  select(where(is.factor) | where(is.character)) %>%
  names()
```

Number of Unique Values:
``` {R, echo=FALSE}
df %>%
  summarise(across(all_of(categorical_cols), n_distinct)) %>%
  pivot_longer(cols = everything(), names_to = "name", values_to = "n") %>%
  pwalk(~ cat(sprintf("%s: %d\n", ..1, ..2)))
```

Overview of the data:
``` {r, echo = FALSE}
glimpse(df)
```

This dataset is well suited for our questions because it captures both the primary global market drivers and the local economic buffers that dictate fuel costs. For Q1 (Regression), `brent_crude_usd` provides the baseline upstream cost, while the `date` and `country` variables allow the model to track the exact monthly time lag it takes for global price shocks to pass through to local retail pumps (`petrol_usd_liter`).

For Q2 (Classification), the data provides the structural macro-context needed to categorize quarterly impact risk bands. Features like `subsidy_level` and `tax_percentage` act as policy shields that alter how a nation absorbs financial shocks, while `income_level` and `region` help the classifier identify which clusters of countries are highly vulnerable versus those with the fiscal resilience to smooth out volatile energy spikes.

---

## Data Preprocessing

This process is divided into three main steps: data cleaning, data integration, and data transformation.

### Data Cleaning

Libraries used for data cleaning:
``` {r data-cleaning, echo=FALSE}
library(tidyverse)
```

#### Indetify the Dirty Data

This step checks for missing values, duplicates, data type inconsistencies, outliers and inconsistent naming.

Missing values per column:
``` {r identify-dirty-data}
raw_data <- df
colSums(is.na(raw_data))
```

Number of duplicates:
``` {r check-duplicates}
sum(duplicated(raw_data))
```

Data type inconsistencies:
``` {r check-data-types}
class(raw_data$petrol_usd_liter)
```

Check for outliers:
``` {r check-outliers}
# D. Check for Outliers
summary(raw_data$diesel_usd_liter)
```

Check for inconsistent naming:
``` {r check-inconsistent-naming}
# E. Check for Inconsistent Naming
unique(raw_data$country)
```