---
title: "Global Fuel Price Data Cleaning Report"
date: "2026-06-06"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Executive Summary
This document details the reproducible data cleaning workflow applied to the dirty_fuel_data.csv dataset. The goal of this pipeline is to address duplicate rows, inconsistent categorical naming, character-to-numeric type mismatches, statistical outliers, and missing data imputation.

## Clean Data Overview
Total Observations: 27,468
Total Columns: 10
Total Unique Countries Covered: 84
Petrol Price Range: $0.010 - $6.779 USD/Liter
Diesel Price Range: $0.010 - $6.240 USD/Liter

## Part 1: Environment Setup
First, we load the tidyverse suite (which includes dplyr, readr, stringr, and ggplot2 for efficient data manipulation) and import the raw dataset.

```{r load-data}
# Load the tidyverse core packages
library(tidyverse)

# Load the dirty data
raw_data <- read_csv("dirty_fuel_data.csv")

# Look at the structure of the data
glimpse(raw_data)
```

## Part 2: Identifying "Dirty" Data
Before writing the cleaning pipeline, we must profile the dataset to uncover quality, structure, and validity issues.

### A. Check for Missing Values (NAs)

```{r check-nas}
print("Missing values per column:")
colSums(is.na(raw_data))
```

### B. Check for Duplicates

```{r check-duplicates}
print("Number of duplicate rows:")
sum(duplicated(raw_data))
```

### C. Check for Data Type Mismatches

```{r check-types}
# Verifying if the petrol price column is recognized as numeric
class(raw_data$petrol_usd_liter)
```

### D. Check for Outliers

```{r check-outliers}
# Summarizing diesel to identify mathematically impossible values
summary(raw_data$diesel_usd_liter)
```

### E. Check for Inconsistent Naming

```{r check-naming}
# Identifying naming variants for categorical grouping
unique(raw_data$country)
```

## Part 3: The Data Cleaning Pipeline
Using a single dplyr pipe sequence, we apply five specific data-cleaning steps to ensure the integrity of the data without breaking spatial correlations.
1. Deduplication: Removed identical duplicate rows to prevent skewed statistical distributions.
2. Structural Fixes: Standardized categorical country values (e.g., mapping "USA" to "United States").
3. Type Conversion: Stripped special characters ($, USD, spaces) from petrol_usd_liter and cast it to numeric.
4. Outlier Mitigation: Filtered out unrealistic diesel prices (>50 or <0) and set them to NA.
5. Missing Value Imputation: Imputed missing values using the grouped country mean, preserving localized geographic price variations.

```{r cleaning-pipeline}
cleaned_data <- raw_data %>%
  # 1. Remove Duplicates
  distinct() %>%
  
  # 2. Fix Categorical Errors (Standardize Country Names)
  mutate(country = case_when(
    country == "USA" ~ "United States",
    country == "U.S.A." ~ "United States", # Accommodating common variants
    TRUE ~ country
  )) %>%
  
  # 3. Fix String/Numeric Issues
  mutate(petrol_usd_liter = as.numeric(str_remove_all(petrol_usd_liter, "[\\$, USD]"))) %>%
  
  # 4. Handle Outliers (Coerce bad numbers to NA)
  mutate(diesel_usd_liter = ifelse(diesel_usd_liter > 50 | diesel_usd_liter < 0, NA, diesel_usd_liter)) %>%
  
  # 5. Impute NAs using geographic country averages
  group_by(country) %>%
  mutate(
    petrol_usd_liter = ifelse(is.na(petrol_usd_liter), 
                              mean(petrol_usd_liter, na.rm = TRUE), 
                              petrol_usd_liter),
    diesel_usd_liter = ifelse(is.na(diesel_usd_liter), 
                              mean(diesel_usd_liter, na.rm = TRUE), 
                              diesel_usd_liter)
  ) %>%
  ungroup()
```

## Part 4: Verification & Export
Post-Cleaning Quality Assurance
We run the data profiling checks again to verify that our dataset is now clean.

```{r verification}
# 1. Check for remaining Missing Values
colSums(is.na(cleaned_data))

# 2. Check for remaining Duplicates
sum(duplicated(cleaned_data))

# 3. Verify final data types
class(cleaned_data$petrol_usd_liter)

# 4. Verify distribution ranges without massive outliers
summary(cleaned_data$diesel_usd_liter)
```

## Save Final Dataset
The final, sparkling clean data is saved for analysis and downstream reporting.

```{r export-data}
write_csv(cleaned_data, "fuel_prices_cleaned_final.csv")
print("Cleaning complete! Final data successfully exported.")
```

## Appendix: Data Dictionary
The clean dataset consists of 10 variables outlined below:

```{r appendix-table, echo=FALSE}
# Create a data frame for your data dictionary
appendix_df <- data.frame(
  No = 1:10,
  Variable_Name = c("date", "country", "region", "income_level", "subsidy_level", 
                    "petrol_usd_liter", "diesel_usd_liter", "lpg_usd_liter", 
                    "brent_crude_usd", "tax_percentage"),
  Data_Type = c("Date", "Character", "Character", "Character", "Character", 
                "Numeric", "Numeric", "Numeric", "Numeric", "Numeric"),
  Description = c("The recorded date of fuel prices.",
                  "Name of the country (standardized).",
                  "Geographic region (e.g., North America, Europe).",
                  "Economic classification of the country.",
                  "Level of government fuel subsidies.",
                  "Price of gasoline per liter in USD.",
                  "Price of diesel per liter in USD.",
                  "Price of LPG per liter in USD.",
                  "The global benchmark price for crude oil.",
                  "The percentage of fuel price consisting of tax.")
)

# Render it perfectly as a table
knitr::kable(appendix_df, col.names = c("#", "Variable Name", "Data Type", "Description"))
```