---
title: "Regression Modeling Report"
date: "2026-06-07"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Executive Summary
In this module of our analytics project, we developed an advanced machine learning pipeline to forecast global retail fuel prices. To capture complex, non-linear market behaviours such as regional policy variations and the lag effects of global oil shocks, we executed a comparative tournament between three different methodologies: Linear Regression (Elastic Net), Random Forest, and Extreme Gradient Boosting (XGBoost). 

---

## 2. Import Libraries & Data
In this section, we load the required packages from the tidymodels framework alongside specialized libraries for model explanation (vip) and parallel computation processing engines.

```{r import-data}
library(readr)
library(tidyverse)
library(tidymodels)
library(vip)         # For feature importance plots
library(doParallel)  # For faster parallel processing

# Load the regression dataset
df <- read_csv("data_processed/Q1_Monthly_Regression_Data.csv")
```

---

## 3. Data Preparation & Chronological Sorting
Because this dataset tracks multiple countries across time (panel/time-series structure), random slicing would result in data leakage. We establish an explicit chronological timeline sorting by dates, and encode ordinal classifications scale adjustments.

```{r data-preparation}
# Create a proper Date column and sort chronologically 
df <- df %>%
  mutate(date = as.Date(paste(year, month, "01", sep = "-"))) %>%
  arrange(date)

# Define levels for ordinal categories
tax_levels <- c("Low Tax", "Medium Tax", "High Tax")
income_levels <- c("Low", "Middle", "High")

# Encode categorical data types
df$country    <- as.factor(df$country)
df$region     <- as.factor(df$region)
df$fuel_type  <- as.factor(df$fuel_type)

# Convert ordered factors to numeric scales so ML algorithms can read their progression scale
df$tax_category <- as.numeric(factor(df$tax_category, levels = tax_levels, ordered = TRUE))
df$income_level <- as.numeric(factor(df$income_level, levels = income_levels, ordered = TRUE))
```

---

## 4. Time-Based Data Splitting & Resampling
We split the data sequentially: the oldest 80% is used to build the predictive engine, while the newest 20% remains untouched to simulate real-world forward-testing. Cross-validation uses a rolling-origin time framework to map sliding windows.

```{r time-splitting}
set.seed(123)
# Chronological Split: Oldest 80% for training, newest 20% for testing
data_split <- initial_time_split(df, prop = 0.80)
train_data <- training(data_split)
test_data  <- testing(data_split)

# Calculate how many rows equal exactly "1 Month" of data across all countries
rows_per_month <- train_data %>% 
  count(year, month) %>% 
  summarise(avg_rows = mean(n)) %>% 
  pull(avg_rows) %>% 
  round()

# Set up the rolling origin cross-validation based on rows
# e.g., 24 months lookback window, 6 months evaluation window, sliding 3 months forward
folds <- rolling_origin(
  train_data,
  initial    = rows_per_month * 24, 
  assess     = rows_per_month * 6,  
  skip       = rows_per_month * 3,  
  cumulative = FALSE                
)
```

---

## 5. Recipe & Model Specifications
We define a workflow preprocessing standard to handle new factor variations seamlessly via dummy variable steps. We set up an engine testing matrix across Elastic Net Linear Regressions, XGBoost, and Random Forest specifications.

```{r model-specifications}
base_recipe <- recipe(avg_retail_price ~ ., data = train_data) %>%
  # Assign time/date trackers an "ID" role so they aren't used as direct linear predictors
  update_role(year, month, date, new_role = "ID") %>% 
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

# Linear Model Spec
linear_spec <- linear_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

# XGBoost Spec
xgb_spec <- boost_tree(trees = tune(), tree_depth = tune(), learn_rate = tune()) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Random Forest Spec 
rf_spec <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

# Combine into a Workflow Set
model_workflows <- workflow_set(
  preproc = list(base_rec = base_recipe),
  models = list(linear = linear_spec, xgb = xgb_spec, rf = rf_spec),
  cross = TRUE
)
```

---

## 6. Hyperparameter Tuning (Parallelised)
We run grid search matrices across all 3 structural workflow designs. Multi-core cluster architectures are triggered to decrease model cross-validation training execution runtimes.

```{r grid-tuning}
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)

set.seed(123)
tune_results <- model_workflows %>%
  workflow_map(
    fn = "tune_grid",
    seed = 123,
    resamples = folds,
    grid = 10,
    metrics = metric_set(rmse, mae, rsq, mape, huber_loss)
  )

stopCluster(cl) # Terminate parallel clusters
```

### Resampling Validation Summary Plot
This chart details the cross-validation score summaries across each evaluated grid option.

```{r tuning-summary-plot, fig.width=8, fig.height=5}
autoplot(tune_results, metric = "rmse")
```

---

## 7. Automated Winning Model Extraction
The system reads across the finalised cross-validation profiles, filters down to select the architecture minimising RMSE, and assigns its hyperparameter choices onto the holdout performance verification test set.

```{r winning-model-extraction}
# Programmatically find the absolute best model ID based on the lowest validation RMSE
best_model_id <- tune_results %>%
  rank_results(rank_metric = "rmse", select_best = TRUE) %>%
  filter(rank == 1) %>%
  slice(1) %>%
  pull(wflow_id)

# Extract best tuning results and parameters for that winning model
winning_tune_res <- tune_results %>% extract_workflow_set_result(best_model_id)
best_parameters  <- winning_tune_res %>% select_best(metric = "rmse")

# Finalize the workflow with the winner's best hyperparameters
final_wf <- tune_results %>%
  extract_workflow(best_model_id) %>%
  finalize_workflow(best_parameters)

# Fit to the complete Training data and evaluate on the unseen holdout Test set
final_fit <- last_fit(
  final_wf, 
  data_split, 
  metrics = metric_set(rmse, mae, rsq, mape, huber_loss)
)
```

The algorithm has dynamically selected **`r best_model_id`** as the winning model structure. Below are its optimal tuned parameters:

```{r print-best-parameters}
knitr::kable(best_parameters, caption = "Winning Model Hyperparameter Values")
```

### Unseen Holdout Test Performance Metrics
The table below presents the model's accuracy scores evaluated on the holdout test phase across all 5 requested evaluation metrics:

```{r print-final-metrics}
final_metrics <- collect_metrics(final_fit)
knitr::kable(final_metrics, caption = "Holdout Test Set Metric Diagnostics")
```

---

## 8. Output Generation
### A. Scatterplot: Actual vs. Predicted Prices
This chart verifies goodness-of-fit. Points centering tightly around the 45-degree diagonal path illustrate a robust baseline alignment.

```{r actual-vs-predicted-plot, fig.width=7, fig.height=5}
predictions_df <- collect_predictions(final_fit)

scatterplot <- ggplot(predictions_df, aes(x = avg_retail_price, y = .pred)) +
  geom_point(alpha = 0.4, color = "midnightblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Actual vs. Predicted Retail Prices",
    subtitle = paste("Winning Model Architecture:", best_model_id),
    x = "Actual Retail Price",
    y = "Predicted Retail Price"
  )
print(scatterplot)
```

### B. Residuals Analysis Plot
Residual points scattered randomly around the center line indicate that the predictive errors remain unbiased across the scale.

```{r residuals-analysis-plot, fig.width=7, fig.height=5}
predictions_df <- predictions_df %>% mutate(residual = avg_retail_price - .pred)

residuals_plot <- ggplot(predictions_df, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.4, color = "darkred") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Model Residuals Plot",
    subtitle = "Points should be randomly scattered around the 0 line",
    x = "Predicted Price (Fitted Values)",
    y = "Residual (Actual - Predicted)"
  )
print(residuals_plot)
```

### C. Relative Feature Importance Plot
This chart highlights which specific macroeconomic variations, regional variables, or tax policies dominate the winning model's internal prediction splits.

```{r feature-importance-plot, fig.width=7, fig.height=5}
final_model_fit <- extract_fit_parsnip(final_fit)

vip_plot <- vip(final_model_fit, num_features = 10, geom = "point") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Top 10 Most Important Predictors",
    subtitle = paste("Extracted dynamically from:", best_model_id),
    x = "Predictors/Features",
    y = "Relative Importance Weight"
  )
print(vip_plot)
```