---
title: "Global Fuel Price Forecasting: Regression Evaluation Dashboard"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
```

## 1. Environment Setup

Load the required packages and create the output folders.

```{r environments}
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(yardstick)

# Strict pathing constraints to neutralize working directory mismatches
dir_model_in <- "models/regression/"
dir_eval_out <- "models/evaluation_regression/"

if (!dir.exists(dir_eval_out)) {
  dir.create(dir_eval_out, recursive = TRUE)
}
```

## 2. Load Trained Models and Test Data 

Import the saved models and the exact test dataset snapshot.

```{r ingestion}
# Path-safe validation check before deserializing objects
paths <- c(
  paste0(dir_model_in, "final_linear_model.rds"),
  paste0(dir_model_in, "final_xgboost_model.rds"),
  paste0(dir_model_in, "final_rf_model.rds"),
  paste0(dir_model_in, "test_data_snapshot.rds")
)

for (p in paths) {
  if (!file.exists(p)) stop(paste("Critical file missing at destination path:", p))
}

final_fit_linear <- readRDS(paste0(dir_model_in, "final_linear_model.rds"))
final_fit_xgb <- readRDS(paste0(dir_model_in, "final_xgboost_model.rds"))
final_fit_rf <- readRDS(paste0(dir_model_in, "final_rf_model.rds"))
test_data_cleaned <- readRDS(paste0(dir_model_in, "test_data_snapshot.rds"))
```

## 3. Generate Predictions

Use `augment()` to calculate predictions and residuals for all models.

```{r inference-matrix}
scored_data_long <- bind_rows(
  augment(final_fit_linear, new_data = test_data_cleaned) %>% mutate(Model = "Elastic Net Linear"),
  augment(final_fit_xgb, new_data = test_data_cleaned) %>% mutate(Model = "XGBoost"),
  augment(final_fit_rf, new_data = test_data_cleaned) %>% mutate(Model = "Random Forest")
) %>%
  mutate(Residual = avg_retail_price - .pred)
```

## 4. Model Performance Comparison

Calculate standard error metrics to find the best performing model.

```{r baseline-benchmarking}
# Instantiate metric set tracking continuous error distances
regression_metrics <- metric_set(rmse, mae, rsq, mape, huber_loss)

comparison_df <- scored_data_long %>%
  group_by(Model) %>%
  regression_metrics(truth = avg_retail_price, estimate = .pred) %>%
  select(Model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate) %>%
  inner_join(
    scored_data_long %>%
      group_by(Model) %>%
      summarise(Max_Error = max(abs(Residual), na.rm = TRUE), .groups = "drop"),
    by = "Model"
  )

# Output styled interactive summary table
comparison_df %>%
  kbl(caption = "Predictive Accuracy Comparison Matrix", booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(which.min(comparison_df$rmse), bold = TRUE, background = "#e8f8f5")
```

## 5. Prediction Speed Benchmark

Measure how fast each model runs predictions on the test set.

```{r runtime-profiles}
speed_tracking <- map_df(
  list("Elastic Net Linear" = final_fit_linear, "XGBoost" = final_fit_xgb, "Random Forest" = final_fit_rf),
  function(wf) {
    start_time <- Sys.time()
    invisible(predict(wf, new_data = test_data_cleaned))
    duration <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
    return(data.frame(Execution_Time_Secs = round(duration, 5)))
  },
  .id = "Model"
)

speed_tracking %>%
  kbl(caption = "Operational Latency & Speed Benchmark Profiles") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 6. Performance by Subgroup

Break down error metrics by fuel type and country to check for specific model biases.

```{r deep-dive-diagnostics}
# Product Categorization Group Breakdown
fuel_diagnostic_matrix <- scored_data_long %>%
  group_by(Model, fuel_type) %>%
  regression_metrics(truth = avg_retail_price, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "mae")) %>%
  select(Model, fuel_type, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)

# Geographical Regional Breakdown
regional_diagnostic_matrix <- scored_data_long %>%
  group_by(Model, country) %>%
  regression_metrics(truth = avg_retail_price, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "mae")) %>%
  select(Model, country, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)

fuel_diagnostic_matrix %>%
  kbl(caption = "Subgroup Error Distributions by Fuel Type Classification") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## 7. Visual Dashboards

### Diagram A: Error Comparison

Predictive Error Profile Layouts explicitly quantify how effectively each algorithm minimizes price prediction errors across MAE, RMSE, and R-squared metrics. XGBoost decisively dominates the comparison, dropping prediction errors to a mere fraction of the other models with a remarkably low MAE and RMSE, drastically outperforming Random Forest’s higher error rates. This massive reduction in error is paired with a near-perfect R-squared value of 0.998, confirming that XGBoost captures nearly 100% of the price variance and establishes itself as the most accurate model by every metric.

```{r visual-errors, echo=FALSE}
plot_metrics <- comparison_df %>%
  pivot_longer(cols = c(rmse, mae, rsq), names_to = "Metric", values_to = "Value") %>%
  # Simplify model names for the axis labels
  mutate(Model = case_when(
    Model == "Elastic Net Linear" ~ "LR",
    Model == "Random Forest" ~ "RF",
    TRUE ~ Model
  ))
ggplot(plot_metrics, aes(x = Model, y = Value, fill = Model)) +
  geom_col(alpha = 0.85, color = "black") +
  geom_text(aes(label = sprintf("%.3f", Value)), vjust = -0.5, fontface = "bold", size = 3.5) +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_viridis_d(option = "mako", begin = 0.2, end = 0.8) +
  labs(title = "Predictive Error Profile Layouts", subtitle = "Standardized model performance comparison scale", x = "", y = "") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", strip.text = element_text(face = "bold"), axis.text.x = element_text(face = "bold"))
```

### Diagram B: Actual vs. Predicted Values

The Goodness-of-Fit plots reveal that while Elastic Net and Random Forest systematically underpredict higher values and visibly bow below the 45-degree line once actual prices exceed 4, XGBoost demonstrates near-perfect calibration. Its predictions tightly hug the diagonal parity line across the entire pricing spectrum, confirming its ability to accurately capture fluctuations and peaks with minimal deviation.

```{r visual-fit, echo=FALSE}
ggplot(scored_data_long, aes(x = avg_retail_price, y = .pred)) +
  geom_point(alpha = 0.25, color = "#2c3e50") +
  geom_abline(slope = 1, intercept = 0, color = "#e74c3c", linetype = "dashed", linewidth = 1) +
  facet_wrap(~Model) +
  labs(title = "Goodness-of-Fit Target Mappings", subtitle = "Evaluated against 45-degree real-world baseline parity line", x = "Actual Retail Market Price", y = "Engine Predicted Value") +
  theme_light(base_size = 12) +
  theme(strip.background = element_rect(fill = "#34495e"), strip.text = element_text(color = "white", face = "bold"))
```

### Diagram C: Speed vs. Accuracy Trade-Off

The operational efficiency frontier evaluates model suitability for deployment by analyzing the multi-objective trade-off between predictive accuracy (RMSE) and computational latency. While the Elastic Net Linear model minimizes inference latency at the expense of precision, the Random Forest architecture proves suboptimal, exhibiting both elevated error rates and high computational overhead. Consequently, XGBoost establishes the optimal Pareto-efficient compromise, yielding the lowest overall error rate while maintaining a highly acceptable inference speed of approximately one second.

```{r visual-frontier, echo=FALSE}
frontier_df <- comparison_df %>% inner_join(speed_tracking, by = "Model")

ggplot(frontier_df, aes(x = Execution_Time_Secs, y = rmse, color = Model, label = Model)) +
  geom_point(size = 6, alpha = 0.9) +
  geom_text(vjust = -1.5, fontface = "bold") +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "The Operational Efficiency Frontier Curve", subtitle = "Balancing predictive power against runtime inference latencies", x = "Prediction Inference Speed (Seconds)", y = "Root Mean Squared Error (RMSE)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  expand_limits(
    y = c(min(frontier_df$rmse) * 0.9, max(frontier_df$rmse) * 1.1),
    x = c(min(frontier_df$Execution_Time_Secs) * 0.5, max(frontier_df$Execution_Time_Secs) * 1.5)
  )
```

---

## 8. Export Results

Save all generated data frames to the evaluation folder for reporting.

```{r exporting}
saveRDS(comparison_df, paste0(dir_eval_out, "metrics_comparison.rds"))
saveRDS(speed_tracking, paste0(dir_eval_out, "operational_speed.rds"))
saveRDS(scored_data_long, paste0(dir_eval_out, "error_profiles_snapshot.rds"))
saveRDS(fuel_diagnostic_matrix, paste0(dir_eval_out, "fuel_deep_dive_metrics.rds"))
saveRDS(regional_diagnostic_matrix, paste0(dir_eval_out, "regional_deep_dive_metrics.rds"))
```
